---
title: 'Book Notes: Computer Age Statistical Inference -- Ch9 Survival Analysis'
author: ''
date: '2020-06-13'
slug: book-notes-computer-age-statistical-inference-ch-9-survival-analysis
categories:
  - Book notes
tags:
  - Introduction
  - Survival Analysis
  - Slides
output:
  blogdown::html_page:
    toc: true
    toc_depth: 2
---

***For the pdf slides, click [here](/pdf/052219_computer_age_stat_inference_Part2_Ch9.pdf)***

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE);

#Color Format
col_text = function(color, x){
  outputFormat = knitr::opts_knit$get('rmarkdown.pandoc.to');
  if(outputFormat == 'latex'){
    if(color == 'green'){
      paste('\\textcolor[RGB]{0, 128, 0}{', x, '}', sep = '');
    } else {
      paste('\\textcolor{', color, '}{', x, '}', sep = '');
    }
  } else if(outputFormat == 'beamer'){
    paste('\\', color, '{', x, '}', sep = '');
  } else if(outputFormat %in% c('html', 'slidy')){
    paste("<font color='", color, "'>", x, '</font>', sep = '');
  } else{
    x;
  }
}

blue = function(x){col_text('blue', x);}
red = function(x){col_text('red', x);}
green = function(x){col_text('green', x);}
```

# Survival Analysis

## Life Table and Kaplan-Meier Estimate

### Life table
* An insurance company's `r blue('life table')` shows information of clients by 
their age. For each age $i$, it contains

    - $n_i$: number of clients
    - $y_i$: number of death
    - $\hat{h}_i = y_i / n_i$: hazard rate
    - $\hat{S}_i$: survival probability estimate 

* An example life table

|Age    |$n$  |$y$  |$\hat{h}$  |$\hat{S}$  |      
|-------|-----|-----|-----------|-----------|
|34     |120  |0    |0.000      |1.000      |
|35     |71   |1    |0.014      |0.986      |
|36     |125  |0    |0.000      |0.986      |
|...    |...  |...  |...        |...        |

### Discrete survival analysis: notations    
* A client's lifetime (`r blue('time until event')`): random variable $X$
    - Also called `r blue('failure time')`, `r blue('survival time')`, or 
    `r blue('event time')`

* Probability of dying at age $i$
\[
f_i = P(X = i)
\]

* Probability of surviving past age $i$
\[
S_i = \sum_{j \geq i + 1} f_j = P(X > i)
\]

* `r blue('Hazard rate')` at age $i$: conditional probability
\[
h_i = \frac{f_i}{S_{i-1}} = P(X = i \mid X \geq i)
\]

### Life table estimations

* Hazard rate estimation: binomial proportions
\[
\hat{h}_i = \frac{y_i}{n_i}
\]
    - Typical frequentist inference: probabilistic results $h_i$ is 
    estimated by the plug-in principle

* Probability of surviving past age $j$ given survival past age $i$:
\[
P(X > j \mid X > i) = \prod_{k = i+1}^j P(X > k \mid X \geq k) 
= \prod_{k = i+1}^j (1 - h_k)
\]

* Probability of survival estimation
\[
\hat{S}_j = \prod_{k={i_0}}^j \left( 1 - \hat{h}_k\right)
\]
where $i_0$ is the starting age

### Continuous survival analysis: notations

* `r blue('Time until event')` $T$: a continuous positive random variable, 
with pdf $f(t)$ and cdf $F(t)$

* `r blue('Survival function')` (i.e., reverse cdf)
\[
S(t) = \int_{t}^{\infty} f(x) dx = P(T > t) = 1- F(t)
\]

* `r blue('Hazard rate')`, also called `r blue('hazard function')`
\[
h(t) = \frac{f(t)}{S(t)} = 
\lim_{\Delta t \rightarrow 0} \frac{P(t < T \leq t + \Delta t \mid T > t)}{\Delta t}
\]
    - In some other books, hazard rate is denoted as $\lambda(t)$
    

### Hazard rate and cumulative hazard function

* Connection between hazard rate $h(t)$ and survival function $S(t)$
\[
h(t) = -\frac{\partial \log S(t)}{\partial t} \quad
\Longleftrightarrow \quad
S(t) = \exp\left\{ -\int_0^t h(x)dx \right\}
\]

* `r blue('Cumulative hazard function')`
\[
\Lambda(t) = \int_0^t h(x) dx = -\log S(t)
\]

* `r red('Knowing any of')` $S(t)$, $h(t)$, $\Lambda(t)$ `r red('allows one to derive the other two')`

* `r green('Example: exponential distributed $T$')`
\[
f(t) = \lambda e^{- \lambda t} \quad \Longrightarrow \quad
S(t) = e^{-\lambda t}, \quad h(t) = \lambda
\]
    - Constant hazard rate: menoryless

### Censored data

* `r blue('Censored data')`: survival times known only to exceed the reported value
    - E.g., lost to followup, experiment ended with some patients still alive
    - Usually denoted as "number+"
    
* Observation $z_i$ for censored data: 
\[
z = (t_i, d_i), 
\]
where $t_i$ is the survival time, and $d_i$ is the indicator
\[
d_i = \begin{cases}
      1 & \text{if death observed}\\
      0 & \text{if death not observed}
      \end{cases}
\]

### Kaplan-Meier estimate

* Among the censored data $z_1, \ldots, z_n$, we denote the 
`r blue('ordered survival times')` as
\[
t_{(1)} < t_{(2)} < \ldots < t_{(n)},
\]
assuming no ties.

* The `r blue('Kaplan-Meier estimate')` for survival probability 
$S_{(j)} = P(X > t_{(j)})$ is the life table estimate
\[
\hat{S}_{(j)} = \prod_{k \leq j} \left( \frac{n-k}{n-k+1} \right)^{d_{(k)}}
\]

* Life table curves are nonparametric: no relationship is assumed between
the hazard rates $h_i$

### A parametric approach 

* Death counts $y_k$ are independent Binomials
\[
y_k \stackrel{ind}{\sim} \text{B}(n_k, h_k)
\]

* Logistic regression
\[
log\left( \frac{h_k}{1-h_k} \right) = \boldsymbol\alpha \mathbf{x}_k
\]

    - E.g., cubic regression:
    \[
    x_k = (1, k, k^2, k^3)'
    \]
    
    - E.g., cubic-linear spline: 
    \[
    x_k = (1, k, (k - k_0)_-^2, (k - k_0)_-^3)'
    \]
    where $x_- = x \cdot \mathbf{1}_{x \leq 0}$

## Cox's Proportional Hazards Model

### Cox's proportional hazards model

* `r blue('Proportional hazards model')` assumes
\[
h_i(t) = h_0(t) \cdot e^{\mathbf{x}_i' \boldsymbol\beta},
\]
where $h_0(t)$ is a baseline hazard, which we don't need to specify

* Denote $\theta_i = e^{\mathbf{x}_i' \boldsymbol\beta}$, then
\[
S_i(t) = S_0(t)^{\theta_i},
\]
where $S_0(t)$ is the baseline survival function

    - Larger value of $\theta_i$ indicates more quickly declining (i.e., worse) 
    survival curves
    - Positive value of the coefficient $\beta_j$ indicates increase of
    the corresponding covariate $x_j$ associating with worse survival curves

### Proportional hazards model: key results

* Let $J$ be the number of observed deaths, occurring at times
\[
T_{(1)} < T_{(2)} < \ldots < T_{(J)}
\]
assuming no ties

* Just before time $T_{(j)}$ there is a `r blue('risk set')` of 
individuals still under observation
\[
R_j = \{i, t_i \geq T_{(j)}\}
\]

* `r red('Key results of the proportional hazards model')`: given one person
dies at time $T_{(j)}$, the probablity it is person $i$, among the set of 
people at risk, is
\[
P(i_j = i \mid R_j) = \frac{e^{\mathbf{x}_i' \boldsymbol\beta}}
{\sum_{k \in R_j} e^{\mathbf{x}_j' \boldsymbol\beta}} = 
\frac{\theta_i}{\sum_{k \in R_j} \theta_j}
\]

### Parameter estimation: based on the partial likelihood

* Estimaiton of $\boldsymbol\beta$ is to maximize the 
`r blue('partial likelihood')`
\[
L(\boldsymbol\beta) = \prod_{j=1}^J \frac{e^{\mathbf{x}_{i_j}' \boldsymbol\beta}}
{\sum_{k \in R_j} e^{\mathbf{x}_j' \boldsymbol\beta}}
\]
where individual $i_j$ dies at time $T_{(j)}$

* `r red('Semi-parametric')`: we do not need to specify the baseline $h_0(t)$,
since it is not contained in the objective function 




```{r}
#gfr = read.table("https://web.stanford.edu/~hastie/CASI_files/DATA/gfr.txt")
```


### References

* Efron, Bradley and Hastie, Trevor  (2016), *Computer Age Statistical Inference*. 
  Cambridge University Press


