---
title: 'Book Notes: Intro to Time Series and Forecasting -- Ch6 ARIMA Models'
author: ''
date: '2020-03-21'
slug: book-notes-intro-to-time-series-and-forecasting-ch6-arima-models
categories:
  - Book notes
tags:
  - Time Series
  - Slides
output:
  blogdown::html_page:
    toc: true
    toc_depth: 2
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE);

#Color Format
col_text = function(color, x){
  outputFormat = knitr::opts_knit$get('rmarkdown.pandoc.to');
  if(outputFormat == 'latex'){
    if(color == 'green'){
      paste('\\textcolor[RGB]{0, 128, 0}{', x, '}', sep = '');
    } else {
      paste('\\textcolor{', color, '}{', x, '}', sep = '');
    }
  } else if(outputFormat == 'beamer'){
    paste('\\', color, '{', x, '}', sep = '');
  } else if(outputFormat %in% c('html', 'slidy')){
    paste("<font color='", color, "'>", x, '</font>', sep = '');
  } else{
    x;
  }
}

blue = function(x){col_text('blue', x);}
red = function(x){col_text('red', x);}
green = function(x){col_text('green', x);}
```

***`r red('For the pdf slides, click [here](/pdf/012119_time_series_ch6.pdf)')`***

### When data is not stationary

* Implication of not stationary: sample ACF or sample PACF 
do not rapidly decrease to zero as lag increases

* What shall we do?
    - Differencing, then fit an ARMA $\rightarrow$ ARIMA
    - Transformation, then fit an ARMA
    - Seasonal model $\rightarrow$ SARIMA

### A non-stationary exmaple: Dow Jones utilities index data

```{r, echo = TRUE, message = FALSE, warning = FALSE, fig.height = 3, fig.width = 9}
library(itsmr); ## Load the ITSM-R package
par(mfrow = c(1, 3));
plot.ts(dowj, main = 'Raw data'); 
acf(dowj); pacf(dowj);
```

### After differencing

```{r, echo = TRUE, message = FALSE, warning = FALSE, fig.height = 3, fig.width = 9}
par(mfrow = c(1, 3));
dowj_diff = dowj[-length(dowj)] - dowj[-1];
plot.ts(dowj_diff, main = 'Data after differencing'); 
acf(dowj_diff); pacf(dowj_diff);
```


# ARIMA Models

### ARIMA model: definition

* `r blue('Autoregressive integrated moving-average models (ARIMA)')`:
Let $d \in \mathbb{N}$, then series $\{X_t\}$ is an ARIMA$(p,d,q)$ process if
\[
Y_t = (1 - B)^d X_t
\]
is a **causal** ARMA$(p,q)$ process.

* `r blue('Difference equation (DE)')` for an ARIMA$(p,d,q)$ process
\[
\phi^*(B) X_t = \phi(B) (1-B)^d X_t = \theta(B) Z_t, \quad
\{Z_t\} \sim \textrm{WN}(0, \sigma^2)
\]
    - $\phi(z)$: polynomial of degree $p$, and $\phi(z) \neq 0$ for $|z| \leq 1$
    - $\theta(z)$: polynomial of degree $q$
    - $\phi^*(z) = \phi(z) (1-z)^d$: has a zero of order $d$ at $z=1$
    
* `r red('An ARIMA process with $d > 0$ is NOT stationary!')`
    
### ARIMA mean is not dertermined by the DE

* $\{X_t\}$ is an ARIMA$(p,d,q)$ process. We can add an arbitrary polynomial 
  trend of degree $d-1$
  \[
  W_t = X_t + A_0 + A_1 t + \cdots + A_{d-1} t^{d-1}
  \]
  with $A_0, \ldots, A_{d-1}$ being any random variables, and $\{W_t\}$
  still satisfies the same ARIMA$(p,d,q)$ difference equation
  
* In other words, the ARIMA DE determines the second-order properties of 
  $\{(1-B)^d X_t\}$ but not those of $\{X_t\}$
  
    - For parameter estimation: $\boldsymbol\phi$, $\boldsymbol\theta$, 
      and $\sigma^2$ are estimated based on $\{(1-B)^d X_t\}$ rather than $\{X_t\}$
    - For forecast, we need additional assumptions

### Fit data using ARIMA processes

* Whether to fit a finite time series using 
    - non-stationary models (such as ARIMA), or 
    - directly using stationary models (such as ARMA)?
    
* If the fitted stationary ARMA model's $\phi(\cdot)$ have zeros very close
  to unit circles, then fitting an ARIMA model is better
    - Parameter estimation is stable
    - The differenced series may only need a low-order ARMA

* `r red('Limitation of ARIMA')`: only permits data to be nonstationary in a 
  very special way
    - Non-stationary: can have zeros anywhere on the unit circle $|z|=1$ 
    - ARIMA model: only has a zero of multiplicity $d$ at the point $z=1$

## Transformation and Identification Techniques

### Natural log transformation

* When data variance increases with mean, it's common to apply log
transformation before fitting the data using ARIMA or ARMA.

* `r green('When does log transfomation work well?')` Suppose that 
  \[
  E(X_t) = \mu_t, \quad Var(X_t) = \sigma^2 \mu_t^2
  \]
  Then by first-order Taylor expansion of $\log(X_t)$ at $\mu_t$:
  \[
  \log(X_t)  \approx \log(\mu_t) + \frac{X_t - \mu_t}{\mu_t} \ \Longrightarrow \ 
  Var\left[\log(X_t)\right]  \approx \frac{Var(X_t)}{\mu_t^2} = \sigma^2
  \]
  The data after log transformation $\log(X_t)$ has a constant variance
  
* Note: log transformation can only be applied to positive data
* Note: If $Y_t = \log(X_t)$, then because expectation and logarithm are not
interchangeable,
\[
\hat{X}_t \neq \exp(\hat{Y}_t)
\]
  

### Generalize the log transformation: Box-Cox transformation

* `r blue('Box-Cox transformation')` 
\[
f_{\lambda}(x) = 
\begin{cases}
\frac{x^{\lambda} - 1}{\lambda}, &  x \geq 0, \lambda > 0 \\
\log(x), &  x > 0, \lambda = 0  \\
\end{cases}
\]

    - Usual range: $0 \leq \lambda \leq 1.5$
    - Common values: $\lambda = 0, 0.5$ 

* Note: $\lim_{\lambda \rightarrow 0} f_{\lambda}(x) = \log(x)$


* Box-Cox transformation can only be applied to non-negative data

## Unit Root Test

### Unit root test for AR$(1)$ process
* $\{X_t\}$ is an AR$(1)$ process: $X_t - \mu = \phi_1(X_{t-1} - \mu) + Z_t$

* Equivalent DE: 
\[
\nabla X_t = X_t - X_{t-1} = \phi_0^* + \phi_1^* X_{t-1} + Z_t
\]
    - where $\phi_0^* = \mu(1 - \phi_1)$ and $\phi_1^* = \phi_1 - 1$
    - Regressing $\nabla X_t$ onto $1$ and $X_{t-1}$, we get the OLS estimator 
      $\hat{\phi}_1^*$ and its standard error $SE(\hat{\phi}_1^*)$

* `r blue('Augmented Dickey-Fuller test for AR$(1)$')` 
    - Hypotheses: $H_0: \phi_1 = 1 \ \longleftrightarrow \  H_1: \phi_1 < 1$
    - Equivalent hypotheses: 
      $H_0: \phi_1^* = 0 \ \longleftrightarrow \ H_1: \phi_1^* < 0$
    - Test statistic: limit distribution under $H_0$ is not normal or t 
      \[
      \hat{\tau} = \frac{\hat{\phi}_1^*}{SE(\hat{\phi}_1^*)}
      \]
    - Rejection region: reject $H_0$ if
      \[
      \begin{cases}
      \hat{\tau} < -2.57 & (90\%)\\
      \hat{\tau} < -2.86 & (95\%)\\
      \hat{\tau} < -3.43 & (99\%)
      \end{cases}
      \]
      
### Unit root test for AR$(p)$ process
* AR$(p)$ process: 
  $X_t - \mu = \phi_1(X_{t-1} - \mu) + \cdots + \phi_p(X_{t-p} - \mu) + Z_t$

* Equivalent DE: 
\[
\nabla X_t = \phi_0^* + \phi_1^* X_{t-1} + \phi_2^* \nabla X_{t-1} + \cdots + \phi_p^* \nabla X_{t-p+1} + Z_t
\]

    - where $\phi_0^* = \mu(1 - \sum_{i=1}^p \phi_i)$, 
      $\phi_1^* = \sum_{i=1}^p \phi_i - 1$, and 
      $\phi_j^* = -\sum_{i=j}^p \phi_i$ for $j \geq 2$
    - Regressing $\nabla X_t$ onto $1, X_{t-1}, \nabla X_{t-1}, \cdots, 
      \nabla X_{t-p+1}$, we get the OLS estimator 
      $\hat{\phi}_1^*$ and its standard error $SE(\hat{\phi}_1^*)$

* `r blue('Augmented Dickey-Fuller test for AR$(p)$')` 
    - Hypotheses: 
      $H_0: \phi_1^* = 0 \ \longleftrightarrow \ H_1: \phi_1^* < 0$
    - Test statistic: 
      \[
      \hat{\tau} = \frac{\hat{\phi}_1^*}{SE(\hat{\phi}_1^*)}
      \]
    - Rejection region: same as augmented Dickey-Fuller test for AR$(1)$
    
### Implement augmented Dickey-Fuller test in R

```{r, echo = TRUE, message = FALSE, warning = FALSE}
library(tseries);
## Note: the lag k here is the AR order p
adf.test(dowj, k = 2);
```


## Forecast ARIMA models

### Forecast an ARIMA$(p, 1, q)$ process

* $\{X_t\}$ is ARIMA$(p, 1, q)$, and $\{Y_t = \nabla X_t\}$ is a causal
ARMA$(p, q)$
\[
X_t = X_0 + \sum_{j=1}^t Y_j, \quad t = 1, 2, \ldots
\]

* Best linear predictor of $X_{n+1}$ 
\[
P_n X_{n+1} = P_n(X_0 + Y_1 + \cdots + Y_{n+1}) = P_n(X_n + Y_{n+1}) 
= X_n + P_n(Y_{n+1}),
\]
    - $P_n$ means based on $\{1, X_0, X_1, \ldots, X_n\}$, or equivalently, 
    $\{1, X_0, Y_1, \ldots, Y_n\}$
    - To find $P_n(Y_{n+1})$, we need to know $E(X_0^2)$ and 
    $E(X_0 Y_j)$, for $j = 1, \ldots, n+1$.
    
* `r red('A sufficient assumption')` for $P_n(Y_{n+1})$ to be the best linear 
   predictor in term of $\{Y_1, \ldots, Y_n\}$:
   $X_0$ is uncorrelated with $Y_1, Y_2, \ldots$

### Forecast an ARIMA$(p, d, q)$ process

* The observed ARIMA$(p, d, q)$ process $\{X_t\}$ satisfies
\[
Y_t = (1-B)^d X_t, \quad t = 1, 2, \ldots, \quad \{Y_t\} \sim \text{causal ARMA}(p, q)
\]

* `r red('Assumption')`: the random vector $(X_{1-d}, \ldots, X_0)$ is 
  uncorrelated with $Y_t$ for all $t > 0$
  
* One-step predictors $\hat{Y}_{n+1} = P_n Y_{n+1}$ and
  $\hat{X}_{n+1} = P_n X_{n+1}$:
  \[
  X_{n+1} - \hat{X}_{n+1} = Y_{n+1} - \hat{Y}_{n+1}
  \]
  
* Recall: the $h$-step predictor of ARMA$(p,q)$ for $n > \max(p, q)$:
\[
P_n Y_{n+h} = \sum_{i=1}^p \phi_i P_n Y_{n+h-i} + 
\sum_{j=h}^q \theta_{n+h-1, j}(Y_{n+h-j} - \hat{Y}_{n+h-j})
\]

* $h$-step predictor of ARIMA$(p, d, q)$ for $n > \max(p, q)$:
\[
P_n X_{n+h} = \sum_{i=1}^{p+d} \phi_i^* P_n X_{n+h-i} + 
\sum_{j=h}^q \theta_{n+h-1, j}(X_{n+h-j} - \hat{X}_{n+h-j})
\]
where $\phi^*(z) = (1-z)^d \phi(z) = 1 - \phi_1^*z - \cdots -\phi_{p+d}^* z^{p+d}$

# Seasonal ARIMA Models

### Seasonal ARIMA (SARIMA) Model: definition
* Suppose $d, D$ are non-negative integers. $\{X_t\}$ is a 
`r blue('seasonal ARIMA$(p, d, q)$')` $\times$ `r blue('$(P, D, Q)_s$
process with period $s$')` if the differenced series
\[
Y_t = (1-B)^d (1-B^s)^D X_t
\]
is a causal ARMA process defined by
\[
\phi(B) \Phi(B^s) Y_t = \theta(B) \Theta(B^s) Z_t, \quad 
\{Z_t\} \sim \textrm{WN}(0, \sigma^2) 
\]
where
\begin{align*}
\phi(z) & = 1 - \phi_1 z - \cdots - \phi_p z^p, \quad
\Phi(z)  = 1 - \Phi_1 z - \cdots - \Phi_P z^P \\
\theta(z) & = 1 + \theta z + \cdots + \theta_q z^q, \quad
\Theta(z)  = 1 + \Theta z + \cdots + \Theta_Q z^Q
\end{align*}

* $\{Y_t\}$ is causal if and only if neither $\phi(z)$ or $\Phi(z)$ has zeros
inside the unit circle

* Usually, $s=12$ for monthly data

### Special case: seasonal ARMA (SARMA)

* `r blue('Between-year model')`: for monthly data,
  each one of the 12 time series is generated by the same
  ARMA$(P, Q)$ model
  \[
  \Phi(B^{12}) Y_t = \Theta(B^{12}) U_t, \quad \{U_{j+12t}, t \in \mathbb{Z}\}
  \sim \textrm{WN}(0, \sigma^2_U)
  \]
  
* `r blue('SARMA$(P, Q)$ with period $s$')`: 
  in the above between-year model, the period 12
  can be changed to any positive integer $s$
    - If $\{U_{t}, t \in \mathbb{Z}\}\sim \textrm{WN}(0, \sigma^2_U)$, then
      the ACVF $\gamma(h) = 0$ unless $h$ divides $s$ evenly. 
      But this may not be ideal for real life application! 
      E.g., this Feb is correlated with last Feb, but not this Jan.
      
* `r blue('General SARMA$(p, q)$')` $\times$ `r blue('$(P, Q)$ with period $s$')`: 
  incorporate dependency between the 12 series by letting $\{U_t\}$ be ARMA:
  \[
  \Phi(B^{s}) Y_t = \Theta(B^{s}) U_t, \quad 
  \phi(B) U_t = \theta(B)Z_t, \quad 
  \{Z_t\} \sim \textrm{WN}(0, \sigma^2)
  \]
    - Equivalent DE for the general SARMA:
    \[
    \phi(B) \Phi(B^s) Y_t = \theta(B) \Theta(B^s) Z_t
    \]
    
### Fit a SARIMA Model

* Period $s$ is known

1. Find $d$ and $D$ to make the difference series $\{Y_t\}$ to look stationary

2. Examine the sample ACF and sample PACF of $\{Y_t\}$ at lags being
   multiples of $s$, to find orders $P, Q$
   
3. Use $\hat{\rho}(1), \ldots, \hat{\rho}(s-1)$ to find orders $p, q$

4. Use AICC to decide among competing order choices

5. Given orders $(p, d, q, P, D, Q)$, estimate MLE of parameters $(\phi, \theta, 
   \Phi, \Theta, \sigma^2)$
 
# Regression with ARMA Errors

### Regression with ARMA errors: OLS estimation

* Linear model with ARMA errors $\mathbf{W} = (W_1, \ldots, W_n)'$:
  \[
  Y_t = \mathbf{x}_t' \boldsymbol\beta + W_t, \quad t = 1, \ldots, n, \quad
  \{W_t\} \sim \text{causal ARMA}(p,q)
  \]
    - Note: each row is indexed by a different time $t$!
    - Error covariance $\boldsymbol\Gamma_n = E(\mathbf{W} \mathbf{W}')$
    
* `r blue('Ordinary least squares (OLS) estimator')`
\[
\hat{\boldsymbol\beta}_{\text{OLS}} =
(\mathbf{X}'\mathbf{X})^{-1}\mathbf{X}'\mathbf{Y}
\]
    - Estimated by minimizing 
    $(\mathbf{Y} - \mathbf{X}\boldsymbol\beta)'
    (\mathbf{Y} - \mathbf{X}\boldsymbol\beta)$
    - `r red('OLS is unbiased')`, even when errors are dependent!

### Regression with ARMA errors: GLS estimation

* `r blue('Generalized least squares (GLS) estimator')`
\[
\hat{\boldsymbol\beta}_{\text{GLS}} =
(\mathbf{X}'\boldsymbol\Gamma_n^{-1}\mathbf{X})^{-1}\mathbf{X}'\boldsymbol\Gamma_n^{-1}\mathbf{Y}
\]
    - Estimated by minimizing the weighted sum of squares
    \[
    (\mathbf{Y} - \mathbf{X}\boldsymbol\beta)' \boldsymbol\Gamma_n^{-1}
    (\mathbf{Y} - \mathbf{X}\boldsymbol\beta)
    \]
    
    - Covariance
      \[
      \textrm{Cov}\left( \hat{\boldsymbol\beta}_{\text{GLS}} \right) = 
      (\mathbf{X}'\boldsymbol\Gamma_n^{-1}\mathbf{X})^{-1}
      \]

    - `r red('GLS is the best linear unbiased estimator')`, i.e., for any
      vector $\mathbf{c}$ and any unbiased estimator $\hat{\boldsymbol\beta}$, 
      we have
      \[
      \textrm{Var}(\mathbf{c}' \hat{\boldsymbol\beta}_{\text{GLS}}) \leq
      \textrm{Var}(\mathbf{c}' \hat{\boldsymbol\beta})
      \]
      
### When $\{W_t\}$ is an AR$(p)$ process

* We can apply $\phi(B)$ to each side of the regression equation and get
  uncorrelated, zero-mean, constant-variance errors
  \[
  \phi(B) \mathbf{Y} = \phi(B) \mathbf{X} \boldsymbol\beta + \phi(B) \mathbf{W} = \phi(B) \mathbf{X} \boldsymbol\beta + \mathbf{Z}
  \]
  
* Under the transformed target variable
  \[
  Y_t^* = \phi(B) Y_t, \quad t = p+1, \ldots, n
  \]
  and the transformed design matrix 
  \[
  X_{t, j}^* = \phi(B) X_{t, j}, \quad t = p+1, \ldots, n, \quad j = 1, \ldots, k
  \]
  the OLS estimator is the best linear unbiased estimator

* Note: after the transformation, the regression sample size reduces to $n-p$

### Regression with ARMA errors: MLE

* MLE of $\boldsymbol\beta, \boldsymbol\phi, \boldsymbol\theta, \sigma^2$ 
  can be estimated by maximizing the Gaussian likelihood with error 
  covariance $\boldsymbol\Gamma_n$

* An iterative scheme

    1. Compute $\hat{\boldsymbol\beta}_{\text{OLS}}$ and regression residuals
        $\mathbf{Y} - \mathbf{X}\hat{\boldsymbol\beta}_{\text{OLS}}$
        
    2. Based on the estimated residuals, compute MLE of the ARMA$(p, q)$ parameters
    
    3. Based on the fitted ARMA model, compute $\hat{\boldsymbol\beta}_{\text{GLS}}$
    
    4. Compute regression residuals 
    $\mathbf{Y} - \mathbf{X}\hat{\boldsymbol\beta}_{\text{GLS}}$, and return to 
    Step 2 until estimators stabilize

* Asymptotic properties of MLE: 
If $\{W_t\}$ is a causal and invertible ARMA, then
    - MLEs are asymptotically normal
    - Estimated regression coefficients are asymptotically independent of 
      estimated ARMA parameters



### References

* Brockwell, Peter J. and Davis, Richard A. (2016), *Introduction to Time Series and Forecasting, Third Edition*. New York: Springer

* Weigt, George (2018) *ITSM-R Reference Manual* http://www.eigenmath.org/itsmr-refman.pdf

* R package `tseries`
https://cran.r-project.org/web/packages/tseries/index.html
