---
title: 'Book Notes: Flexible Imputation of Missing Data -- Ch2 Multiple Imputation'
author: ''
date: '2020-08-23'
slug: book-notes-flexible-imputation-of-missing-data-ch2-multiple-imputation
categories:
  - Book notes
tags:
  - Missing data
  - Introduction
  - Slides
output:
  blogdown::html_page:
    toc: true
    toc_depth: 2
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE);

#Color Format
col_text = function(color, x){
  outputFormat = knitr::opts_knit$get('rmarkdown.pandoc.to');
  if(outputFormat == 'latex'){
    if(color == 'green'){
      paste('\\textcolor[RGB]{0, 128, 0}{', x, '}', sep = '');
    } else {
      paste('\\textcolor{', color, '}{', x, '}', sep = '');
    }
  } else if(outputFormat == 'beamer'){
    paste('\\', color, '{', x, '}', sep = '');
  } else if(outputFormat %in% c('html', 'slidy')){
    paste("<font color='", color, "'>", x, '</font>', sep = '');
  } else{
    x;
  }
}

blue = function(x){col_text('blue', x);}
red = function(x){col_text('red', x);}
green = function(x){col_text('green', x);}
```

***`r red('For the pdf slides, click [here](/pdf/081120_imputation_book_ch2.pdf)')`***

# Concepts in Incomplete Data

### Notations

* $m$: number of multiple imputations
* $Y$: data of the sample 
    - Includes both covariates and response
    - Dimension $n \times p$
* $R$: observation indicator matrix, known
    - A $n \times p$ 0-1 matrix 
    - $r_{ij} =0$ for missing and 1 for observed

* $Y_{\text{obs}}$: observed data
* $Y_{\text{mis}}$: missing data
* $Y = (Y_{\text{obs}}, Y_{\text{mis}})$: complete data

* $\psi$: the parameter for the missing mechanism
* $\theta$: the parameter for the full data $Y$


### Concepts of MCAR, MAR, and MNAR, with notations

* `r blue('Missing completely at random (MCAR)')`
\[
P(R = 0 \mid Y_{\text{obs}}, Y_{\text{mis}}, \psi) = P(R = 0\mid \psi)
\]

* `r blue('Missing at random (MAR)')`
\[
P(R = 0 \mid Y_{\text{obs}}, Y_{\text{mis}}, \psi) = P(R = 0\mid Y_{\text{obs}}, \psi)
\]
    
* `r blue('Missing not at random (MNAR)')`
\[
P(R = 0 \mid Y_{\text{obs}}, Y_{\text{mis}}, \psi) \text{ does not simplify}
\]

### `r blue('Ignorable')`

* The missing data mechanism is `r blue('ignorable')` for likelihood inference (on $\theta$), if

    1. MAR, and
    2. Distinctness: the parameters $\theta$ and $\psi$ are independent (from a Bayesian's view)
    
* If the nonresponse if ignorable, then
\[
P(Y_{\text{mis}} \mid Y_{\text{obs}}, R) = P(Y_{\text{mis}} \mid Y_{\text{obs}})
\]
Thus, if the missing data model is ignorable, we can model $\theta$ just using the observed data

# Why and When Multiple Imputation Works

### Goal of multiple imputation

* `r red('Note: for most multiple imputation practice, this goal is to train a (predictive) model with as small variances of the parameters as possible')`

* $Q$: estimand (the parameter to be estimated)
* $\hat{Q}$: estimate
    - `r green('Unbias')`
    \[
    E(\hat{Q} \mid Y) = Q
    \]
    - `r green('Confidence valid')`: 
    \[
    E(U \mid Y) \geq V(\hat{Q}\mid Y)
    \]
    where $U$ is the estimated covariance matrix of $\hat{Q}$, the expectation is over all possible samples,
    and $V(\hat{Q}\mid Y)$ is the variance caused by the sampling process
    
### Within-variance and between-variance

\begin{align*}
E(Q \mid Y_{\text{obs}}) &= E_{Y_{\text{mis}} \mid Y_{\text{obs}}}\{
  E(Q \mid Y_{\text{obs}}, Y_{\text{mis}})\}\\
V(Q \mid Y_{\text{obs}}) &= \underbrace{E_{Y_{\text{mis}} \mid Y_{\text{obs}}}\{
  V(Q \mid Y_{\text{obs}}, Y_{\text{mis}})\}}_{\text{within-variance}} +
  \underbrace{V_{Y_{\text{mis}} \mid Y_{\text{obs}}}\{
  E(Q \mid Y_{\text{obs}}, Y_{\text{mis}})\}}_{\text{between variance}}
\end{align*}

* Within-variance: average of the repeated complete-data posterior variance of $Q$, estimated by
\[
\bar{U} = \frac{1}{m} \sum_{l=1}^m \bar{U}_l,
\]
where $\bar{U}_l$ is the variance of $\hat{Q}_l$ in the $l$th imputation

* Between-variance: variance between the complete-data posterior means of $Q$, estimated by
\[
B = \frac{1}{m-1}\sum_{l=1}^m\left(\hat{Q}_l - \bar{Q}\right)\left(\hat{Q}_l - \bar{Q}\right)', \quad \bar{Q} = \frac{1}{m} \sum_{l=1}^m \hat{Q}_l
\]

### Decomposition of total variation

* Since $\bar{Q}$ is estimated using finite $m$, the contribution to the variance is about $B/m$. Thus, `r red('the total posterior variance of $Q$ can be decomposed into three parts:')`
\[
T = \bar{U} + B + B/m = \bar{U} + \left(1 + \frac{1}{m}\right) B 
\]

* $\bar{U}$: the conventional variance, due to sampling rather than getting the entire population. 

* $B$: the extra variance due to missing values

* $B/m$: the extra simulation variance because $\bar{Q}$ is estimated for finite $m$

    - Traditionally choices are $m = 3, 5, 10$, but the current advice is to use a larger $m$, e.g., $m = 50$
    
### Properness of an imputation procedure

* An imputation procedure is `r blue('confidence proper')` for complete-data statistics 
$\hat{Q}, U$, if it satisfies the following three conditions approximately at large $m$
\begin{align*}
E\left(\bar{Q} \mid Y\right) & = \hat{Q}\\
E\left(\bar{U} \mid Y\right) & = U\\
\left(1 + \frac{1}{m}\right) E(B \mid Y) & \geq V(\bar{Q})
\end{align*}

    - Here $\hat{Q}$ is the complete-sample estimator of $Q$, and $U$ is its covariance
    - If we replace the $\geq$ by $>$ in the third formula, 
    then the procedure is said to be `r blue('proper')`
    - It is not always easy to check whether a procedure is proper.
    
### Scope of the imputation model

* `r green('Broad')`: one set of imputations to be used for all projects and analyses

* `r green('Intermediate')`: one set of imputations per project and use this for all analyses

* `r green('Narrow')`: a separate imputed dataset is created for each analysis

* Which one is better: depends on the use case

### Variance ratios

* Proportion of variation attributable to the missing data
\[
\lambda = \frac{B + B/m}{T}
\]
    - If $\lambda > 0.5$, then the influence of the imputation model on the final result is larger than that of the complete-data model
    
* `r blue('Relative increase in variance due to nonresponse')`
\[
r = \frac{B + B/m}{\bar{U}} = \frac{\lambda}{1-\lambda}
\]


* `r blue('Fraction of information about $Q$ missing due to nonresponse')`
\[
\gamma = \frac{r + 2/(\nu + 3)}{1 + r} = \frac{\nu + 1}{\nu + 3}\lambda + \frac{2}{\nu + 3}
\]
    - Here, $\nu$ is the degrees of freedom (see next)
    - When $\nu$ is large, $\gamma$ is very close to $\lambda$

### Degrees of freedom (df)

* The degrees of freedom is the number of observations after accounting for the number of parameters in the model.

* The "old" formula (as in Rubin 1987): may produce values larger than the sample size in the complete data
\[
\nu_{\text{old}} = (m-1) \left(1 + \frac{1}{r^2}\right) = \frac{m-1}{\lambda^2}
\]

* Let $\nu_\text{com}$ be the conventional df in a complete-data inference problem. If the number of parameters in the model is $k$ and the sample size is $n$, then $\nu_\text{com} = n-k$. The estimated observed data df that accounts for the missing information is
\[
\nu_{\text{obs}} = \frac{\nu_\text{com} + 1}{\nu_\text{com} + 3} \nu_\text{com} (1-\lambda)
\]

* Barnard-Rubin correction: the adjusted df to be used for testing in multiple imputation is
\[
\nu = \frac{\nu_{\text{old}} \nu_{\text{obs}}}{\nu_{\text{old}} + \nu_{\text{obs}}}
\]

### A numerical example

```{r, echo = FALSE}
library(pander)
panderOptions('round', 3)
```

```{r, echo = TRUE, message = FALSE, warning = FALSE, fig.height = 3, fig.width = 9}
## Load the mice package
library(mice); 
imp <- mice(nhanes, print = FALSE, m = 10, seed = 24415)
fit <- with(imp, lm(bmi ~ age))
est <- pool(fit); print(est, digits = 2)
```

* Columns `ubar`, `b`, and `t` are the variances
* Column `dfcom` is $\nu_\text{com}$
* Column `df` is the Barnard-Rubin correction $\nu$

### T-test for regression coefficients

* Use the Barnard-Rubin correction of $\nu$ as the shape parameter of t-distribution.

```{r, echo = TRUE, message = FALSE, warning = FALSE, fig.height = 3, fig.width = 9}
print(summary(est, conf.int = TRUE), digits = 1)
```

# More about Imputation Methods

### Imputation evaluation criteria

* The following criteria are useful in simulation studies (when you know the true $Q$)

1. Raw bias (RB): upper limit $5\%$
\[
\text{RB} = \left|\frac{E\left(\bar{Q}\right) - Q}{Q}\right|
\]

2. Coverage rate (CR): A CR below $90\%$ for the nominal $95\%$ interval is bad

3. Average width (AW) of confidence interval

4. Root mean squared error (RMSE): the smaller the better
\[
\text{RMSE} = \sqrt{\left(E\left(\bar{Q}\right) - Q\right)^2}
\]

### Imputation is not prediction

* Shall we evaluate an imputation method by examine how it can closely recover the missing values?
    - For example, using the RMSE to see if the imputed values $\dot{y}_i$ are close to the true (removed) 
    missing data $y_i^{\text{mis}}$?
    \[
    \text{RMSE} = \sqrt{\frac{1}{n_{\text{mis}}}\sum_{i=1}^{n_{\text{mis}}}\left(y_i^{\text{mis}} - \dot{y}_i \right)^2}
    \]

* `r red('NO! This will favor least squares estimates, and it will find the same values over and over; and thus it is single imputation. This ignores the inherent uncertainty of the missing values.')`

### When not to use multiple imputation

* For predictive modeling, if the missing values are in the target variable $Y$, then complete-case analysis and multiple imputation are equivalent. 

* Two special cases where listwise deletion is better than multiple imputation

1. If the probability to be missing does not depend on $Y$

2. If the complete data model is logistic regression, and the missing data are confined to $Y$, not $X$



### References

* Van Buuren, S. (2018). Flexible Imputation of Missing Data, 2nd Edition. CRC press.

    - https://stefvanbuuren.name/fimd/

* Rubin, D. (1987). Multiple Imputation for Nonresponse in Surveys. New York: John Wiley & Sons.
